# Developer notes

## Release process

Check spelling.

```{r}
devtools::spell_check()
```

Update documentation and run tests.

```{r}
devtools::document()
devtools::check(vignettes = FALSE)
testthat::test_local()
```

Run R CMD check.

```{r}
pkg_path <- devtools::build(vignettes = TRUE, manual = TRUE)
system(paste("R CMD check", pkg_path))
```

Check for Windows compatibility.

```{r}
devtools::check_win_devel()
```

Build github_document version of the vignette. This is for viewing the vignette
at Github and so does not affect the package build. The resulting updates should
be pushed to the `develop` branch.

```{r}
# Render to GitHub-flavored Markdown
rmarkdown::render("vignettes/briefmanual.Rmd",
                  output_format = "rmarkdown::github_document",
                  output_file = "briefmanual.md")

# Move the markdown files to the root directory
file.rename("vignettes/briefmanual.md", "briefmanual.md")
file.copy(
  list.files("vignettes/briefmanual_files/", full.names = TRUE),
  "briefmanual_files/",
  recursive = TRUE)
unlink("vignettes/briefmanual_files/", recursive = TRUE)
```

Release the package to CRAN. When successful, merge `develop` into `main`.

```{r}
usethis::use_cran_comments()
devtools::release()
```


## Other

Install the local package.

```{r}
devtools::install()
```
